---
title: "NOAA Storm Database: analysis of Impacts on Health and Economy"
author: "Flavio Massimo Saldicco"
date: "Tuesday, February 10, 2015"
output: html_document
---

[]
##Synopsis
tbd

##Data Processing

###Getting the Data File
The NOAA Storm Database contains a record of the events concerning severe weather occurred in the USA. It can be consulted and downloaded in CVS format at <http://www.ncdc.noaa.gov/stormevents/>. It is kept updated by NOAA and currently contains events up to Oct. 2014.

The specific version used for this analysis contains events occurred in the USA from Jan. 1950 to Nov. 2011. It can be downloaded in a BZ2 compressed CSV file at the Coursera Reproducible Research link <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>. This analysis assumes that the data file has already been downloaded and is present in a locally accessible computer folder.

###Downloading the Data File
```{r, echo=TRUE}
#try if file is present in the current home
#if not, download it
datafile <- "repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists(paste0("~/", datafile))) {
    url <- "http://d396qusza40orc.cloudfront.net/"
    download.file(paste0(url, datafile), paste0("~/", datafile))
}
```

###Creating the Dataset
The dataset **weather_events** is created by reading the datafile. The bzfile connection is used, because the original file is BZ2 compressed.
```{r, echo=TRUE}
weather_events <- read.csv(bzfile(paste0("~/", datafile)))
```

###Cleaning the dataset
#convert the event type field (EVTYPE) to upper case
```{r, echo=TRUE}
weather_events["EVTYPE"] <- toupper(weather_events[c("EVTYPE")])
```




##Results
```{r, echo=TRUE}


CasualtiesByEvent <- aggregate(weather_events[c("FATALITIES", "INJURIES")],
                               by=weather_events["EVTYPE"],
                               FUN="sum")


```

weather_events["INJURIES"],











