---
title: "NOAA Storm Database: analysis of Impacts on Health and Economy"
author: "Flavio Massimo Saldicco"
date: "Tuesday, February 10, 2015"
output: html_document
---

[]
##Synopsis
tbd

##Data Processing

###Getting the Data File
The NOAA Storm Database contains a record of the events concerning severe weather occurred in the USA. It can be consulted and downloaded in CVS format at <http://www.ncdc.noaa.gov/stormevents/>. It is kept updated by NOAA and currently contains events up to Oct. 2014.

The specific version used for this analysis contains events occurred in the USA from Jan. 1950 to Nov. 2011. It can be downloaded in a BZ2 compressed CSV file at the Coursera Reproducible Research link <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>. This analysis assumes that the data file has already been downloaded and is present in a locally accessible computer folder.

###Downloading the Data File
```{r, echo=TRUE}
#try if file is present in the current home
#if not, download it
datafile <- "repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists(paste0("~/", datafile))) {
    url <- "http://d396qusza40orc.cloudfront.net/"
    download.file(paste0(url, datafile), paste0("~/", datafile))
}
```

###Creating the Dataset
The dataset **weather_events** is created by reading the datafile. The bzfile connection is used, because the original file is BZ2 compressed.
```{r, echo=TRUE}
weather_events <- read.csv(bzfile(paste0("~/", datafile)))
```

###Cleaning the dataset

```{r, echo=TRUE}

CasualtiesByEvent <- aggregate(weather_events[c("FATALITIES", "INJURIES")],
                               by=weather_events["EVTYPE"],
                               FUN="sum")

head(CasualtiesByEvent, n = 20L)
```

weather_events$EVTYPE <- toupper(as.character(weather_events$EVTYPE))

weather_events$EVTYPE[grepl("AVALAN", weather_events$EVTYPE)] <- "AVALANCHE"
#weather_events$EVTYPE[grepl("COAST|CST", weather_events$EVTYPE)] <- "COASTAL FLOODING"
weather_events$EVTYPE[grepl("COLD", weather_events$EVTYPE)] <- "COLD"
weather_events$EVTYPE[grepl("DRY", weather_events$EVTYPE)] <- "DRY"
weather_events$EVTYPE[grepl("DUST|SMOKE|FUNNEL", weather_events$EVTYPE)] <- "DUST"
weather_events$EVTYPE[grepl("FIRE", weather_events$EVTYPE)] <- "WILDFIRE"
weather_events$EVTYPE[grepl("FLOOD", weather_events$EVTYPE)] <- "FLOOD"
weather_events$EVTYPE[grepl("FREEZ|FROST", weather_events$EVTYPE)] <- "FREEZE"


weather_events$EVTYPE[grepl("HAIL", weather_events$EVTYPE)] <- "HAIL"
weather_events$EVTYPE[grepl("HEAT", weather_events$EVTYPE)] <- "HEAT"
weather_events$EVTYPE[grepl("ICE", weather_events$EVTYPE)] <- "ICE"
weather_events$EVTYPE[grepl("LAND", weather_events$EVTYPE)] <- "LANDSLIDE"

weather_events$EVTYPE[grepl("LIGHT|LIGNT", weather_events$EVTYPE)] <- "LIGHTNING"
weather_events$EVTYPE[grepl("LOW TEMP", weather_events$EVTYPE)] <- "LOW TEMPERATURE"
weather_events$EVTYPE[grepl("MUD", weather_events$EVTYPE)] <- "MUD SLIDE"

weather_events$EVTYPE[grepl("RAIN|PRECIP", weather_events$EVTYPE)] <- "RAIN"
weather_events$EVTYPE[grepl("RIP|ROGUE|ROUGH|DROWN|SURF|SEA|WATER|WAVE|MARINE|TIDE", weather_events$EVTYPE)] <- "ROUGH WATERS"

weather_events$EVTYPE[grepl("HURRI", weather_events$EVTYPE)] <- "HURRICANE"
weather_events$EVTYPE[grepl("SNOW|BLIZZ", weather_events$EVTYPE)] <- "SNOW"
weather_events$EVTYPE[grepl("SPOUT", weather_events$EVTYPE)] <- "WATERSPOUT"
weather_events$EVTYPE[grepl("STORM", weather_events$EVTYPE)] <- "STORM"
weather_events$EVTYPE[grepl("SUMMARY|OTHER|[?]", weather_events$EVTYPE)] <- "UNKNOWN"
weather_events$EVTYPE[grepl("THUNDERSTORM|TSTM", weather_events$EVTYPE)] <- "THUNDERSTORM"
weather_events$EVTYPE[grepl("TORN", weather_events$EVTYPE)] <- "TORNADO"
weather_events$EVTYPE[grepl("UNSEAS|URBAN|RECORD|ABNOR", weather_events$EVTYPE)] <- "OTHER"

weather_events$EVTYPE[grepl("VOLC", weather_events$EVTYPE)] <- "VOLCANO"
weather_events$EVTYPE[grepl("WARM", weather_events$EVTYPE)] <- "WARM"
weather_events$EVTYPE[grepl("WIND|WND|GUST", weather_events$EVTYPE)] <- "WIND"
weather_events$EVTYPE[grepl("WINTER|WINTRY", weather_events$EVTYPE)] <- "WINTER WEATHER"


#factorise EVTYPE
weather_events$EVTYPE <- factor(weather_events$EVTYPE)

According to [], para. 2.7, page 12, only three type of damage exponents should be present: K (1,000), M (1,000,000) and B (1,000,000,000). Unfortunately, the full list of exponents for property damages shows:
unique(weather_events$PROPDMGEXP)

And for crop damages:
unique(weather_events$CROPDMGEXP)

Therefore, we need a strategy to cope with these information.
Let's try to figure out the impact of the unexpected letters in the Damages.

```{r, echo=TRUE}
DamagesByPropExp <- aggregate(weather_events[c("PROPDMG")],
                               by=weather_events["PROPDMGEXP"],
                               FUN="sum")

And in the Crop Damages:

DamagesByCropExp <- aggregate(weather_events[c("CROPDMG")],
                               by=weather_events["CROPDMGEXP"],
                               FUN="sum")
```

Taking into consideration that we don't have a clue concerning the meaning of those letters, and that the impact in economical terms should not substantially modify our analysis, the decision is made to ignore the records corresponding to those letters.
Concerning the lowercase "m" and "k" present, they will be made uppercase and considered as "M" and "K".

```{r, echo=TRUE}
econ_impact <- subset(weather_events, select=c(EVTYPE, PROPDMG, PROPDMGEXP,
    CROPDMG, CROPDMGEXP))
econ_impact$PROPVAL <- 0
econ_impact$CROPVAL <- 0
econ_impact$PROPDMGEXP <- toupper(as.character(econ_impact$PROPDMGEXP))
econ_impact$CROPDMGEXP <- toupper(as.character(econ_impact$CROPDMGEXP))

for (i in 1:nrow(econ_impact)) {

    #properties
    if (econ_impact$PROPDMGEXP[i] == "K") {
        econ_impact$PROPVAL[i] <- econ_impact$PROPDMG[i] * 1000
    } else if (econ_impact$PROPDMGEXP[i] == "M") {
        econ_impact$PROPVAL[i] <- econ_impact$PROPDMG[i] * 1000000
    } else if (econ_impact$PROPDMGEXP[i] == "B") {
        econ_impact$PROPVAL[i] <- econ_impact$PROPDMG[i] * 1000000000
    } else {
        econ_impact$PROPVAL[i] <- 0
    }

    #crop
    if (econ_impact$CROPDMGEXP[i] == "K") {
        econ_impact$CROPVAL[i] <- econ_impact$CROPDMG[i] * 1000
    } else if (econ_impact$CROPDMGEXP[i] == "M") {
        econ_impact$CROPVAL[i] <- econ_impact$CROPDMG[i] * 1000000
    } else if (econ_impact$CROPDMGEXP[i] == "B") {
        econ_impact$CROPVAL[i] <- econ_impact$CROPDMG[i] * 1000000000
    } else {
        econ_impact$CROPVAL[i] <- 0
    }

}
```

##Results
```{r, echo=TRUE}

#impact to health
CasualtiesByEvent <- aggregate(weather_events[c("FATALITIES", "INJURIES")],
                               by=weather_events["EVTYPE"],
                               FUN="sum")


top10Fatalities <- head(CasualtiesByEvent[order(CasualtiesByEvent$FATALITIES, decreasing=TRUE),], 10)
top10Fatalities

top10Injuries <- head(CasualtiesByEvent[order(CasualtiesByEvent$INJURIES, decreasing=TRUE),], 10)

top10Injuries

par(mfrow=c(2,1))
par(cex = 0.6)
barplot(top10Fatalities$FATALITIES,
        names=top10Fatalities$EVTYPE,
        axes=FALSE,
        main="Top 10 Casualties by Type of Event - USA 1950-2011",
        ylab="Fatalities")
axis(2, at=seq(0, 7000, 500))

barplot(top10Injuries$INJURIES,
        names=top10Injuries$EVTYPE,
        axes=FALSE,
        ylab="Injuries",
        sub="Source: NOAA")
axis(2, at=seq(0, 100000, 15000))


```
It appears that tornadoes are, by far, the most dangerous type of event for human health, causing the highest number of both fatalities and injuries.
 

```{r, echo=TRUE}
#impact to economy
DamagesByEvent <- aggregate(econ_impact[c("PROPVAL", "CROPVAL")],
                               by=econ_impact["EVTYPE"],
                               FUN="sum")


top10PropDamages <- head(DamagesByEvent[order(DamagesByEvent$PROPVAL, decreasing=TRUE),], 10)
top10PropDamages

top10CropDamages <- head(DamagesByEvent[order(DamagesByEvent$CROPVAL, decreasing=TRUE),], 10)
top10CropDamages

par(mfrow=c(2,1))
par(cex = 0.6)
barplot(top10PropDamages$PROPVAL,
        names=top10PropDamages$EVTYPE,
        axes=FALSE,
        main="Top 10 Damages by Type of Event - USA 1950-2011",
        ylab="Damages to Properties")
axis(2, at=seq(0, 200e9, 50e9))

barplot(top10CropDamages$CROPVAL,
        names=top10CropDamages$EVTYPE,
        axes=FALSE,
        ylab="Damages to Crop FIelds")
axis(2, at=seq(0, 100000, 15000))




```

##Annex A
List of recorded event types before cleaning
```{r, echo=TRUE}
CasualtiesByEvent["EVTYPE"]
```





